---
title: "Assignment 10: Data Scraping"
author: "Tani Valdez Rivas"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 

# Installing packages
library(tidyverse)
library(lubridate)
library(viridis)
library(here)
library(rvest)

# Checking working directory
here()


```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 

# Using the read_html() to read the contents
NCDEQ_Web <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
NCDEQ_Web



```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 

# Scraping the water system name
the_water_system <- NCDEQ_Web %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
the_water_system

# Scraping the PWSID
the_PWSID <- NCDEQ_Web %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
the_PWSID

# Scraping the ownership
the_ownership <- NCDEQ_Web %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
the_ownership

# Selecting the MGD of the water supply source for each month

max_monthly_MGD <- NCDEQ_Web %>% 
  html_nodes('th~ td+ td') %>%
  html_text()
max_monthly_MGD


```



4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 

# Creating Month and year
Months <- c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")
Year <- 2022

# Using the data.frame function to scrap data into data frame
df_NCDEQ <- data.frame(Month = rep(Months, each = 1),
                       Year = rep(Year, 12),
                       "Monthly_Max_mgd" = as.numeric(max_monthly_MGD))%>% 
  mutate(Water_System = !!the_water_system,
         PWSID = !!the_PWSID,
         Ownership = !!the_ownership,
         Date = my(paste(Month,"-",Year)))

#5 

# Using ggplot to create line plot of dataframe
Max_Withdrawls_2022 <- 
  ggplot(df_NCDEQ,aes(x=Date, y=Monthly_Max_mgd)) + 
  geom_line() + 
  labs(title = paste("2022 Monthly Water usage data for",the_water_system),
       y="Monthly Withdrawal (mgd)",
       x="Date")
Max_Withdrawls_2022
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 

# Using function to scrape any PWSID and year
scrape_NCDEQ <- function(the_PWSID, Year){
  NCDEQ_Web <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', 
                                  the_PWSID, '&year=', Year))

# Using the code from question 3
  
# Scraping the water system name
the_water_system <- NCDEQ_Web %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
the_water_system

# Scraping the PWSID
the_PWSID <- NCDEQ_Web %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
the_PWSID

# Scraping the ownership
the_ownership <- NCDEQ_Web %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
the_ownership

# Selecting the MGD of the water supply source for each month
max_monthly_MGD <- NCDEQ_Web %>% 
  html_nodes('th~ td+ td') %>%
  html_text()
max_monthly_MGD
  
#Convert to a dataframe using data 
df_NCDEQ <- data.frame(Month = rep(Months, each = 1),
                       Year = rep(Year, 12),
                       "Monthly_Max_mgd" = as.numeric(max_monthly_MGD))%>% 
  mutate(Water_System = !!the_water_system,
         PWSID = !!the_PWSID,
         Ownership = !!the_ownership,
         Date = my(paste(Month,"-",Year)))
  

#Return the dataframe
return(df_NCDEQ)
}
```


7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 

# Running the fetch and plot for Durham
Durham2015 <- scrape_NCDEQ('03-32-010',2015)
view(Durham2015)


```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 

# Running the fetch and plot for Ashville
Asheville2015 <- scrape_NCDEQ('01-11-010',2015)
view(Asheville2015)

# Combining Asheville and Durham 2015 data
AshevilleDurham2015 <- bind_rows(Durham2015, Asheville2015)


# Creating ggplot for Durham and Asheville's water withdrawls
AshevilleDurham2015Plot <- 
  ggplot(AshevilleDurham2015, aes(x = Date, y = Monthly_Max_mgd, 
                            color = Water_System)) +
  geom_line() +
  labs(title = "Water Withdrawals of Durham and Asheville (2015)",
       x = "Date",
       y = "Maximum Daily Withdrawals (MGD)")
AshevilleDurham2015Plot
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 

# Creating the sequence through 2010 to 2021
Ashevilles_df_years <- seq(2010,2021)

# Identifying the PWSID
Ashevilles_PWSID = '01-11-010'

#"Map" the scrape function to retrieve data for all these
dfs_Ashevilles_2010_2021 <- map2(Ashevilles_PWSID, Ashevilles_df_years, scrape_NCDEQ) 

#Conflate the returned list of dataframes into a single one
single_df_Asheville_2010_2021<- bind_rows(dfs_Ashevilles_2010_2021)

#Plotting the dataframe
Asheville_2010_2021_plot <-
  ggplot(single_df_Asheville_2010_2021,aes(x = Date, y = Monthly_Max_mgd)) +
  geom_line() +
  scale_x_date(date_breaks="1 year") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
  geom_smooth(method="loess",se=FALSE) +
    labs(title = paste("Asheville's Max Daily Withdrawal from 2010 to 2021"),
         x = "Date",
         y = "Max Daily Withdrawal (MGD)")


Asheville_2010_2021_plot

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
> From 2010 to 2018, the maximum day usage remained constant. But from 2019 and on,
> water usage increased, with the peak occuring between 2019 and 2020. 
