---
title: "Assignment 3: Data Exploration"
author: "Tani Valdez Rivas"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on
Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing
    `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill
    each instruction.
4.  Assign a useful **name to each code chunk** and include ample
    **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code
    into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the
    dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by
manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()`
or `View()` commands exist in your code.

------------------------------------------------------------------------

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse,
    lubridate), and upload two datasets: the ECOTOX neonicotinoid
    dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge
    NEON dataset for litter and woody debris
    (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets
    "Neonics" and "Litter", respectively. Be sure to include the
    subcommand to read strings in as factors.

```{r,  message = FALSE}

# Setting working directory
getwd()

# Installing packages
library(tidyverse) 
library(lubridate)

# Uploading the data sets
Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",
                    stringsAsFactors = TRUE)
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",
                   stringsAsFactors = TRUE)



```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental
    Protection Agency's ECOTOX Knowledgebase, a database for
    ecotoxicology research. Neonicotinoids are a class of insecticides
    used widely in agriculture. The dataset that has been pulled
    includes all studies published on insects. Why might we be
    interested in the ecotoxicology of neonicotinoids on insects? Feel
    free to do a brief internet search if you feel you need more
    background information.

> Answer: Neonicotinoids were first registered for use in the early
> 1990s. It was not until the mid-2000s did the use of neonicotinoid use
> exploded in popularity in the agricultural setting. Although the
> application of insecticide has shifted to seed coating, neoincotinoids
> have a strong effect on non-target organisms. Pollinators and aquatic
> insects have been exposed to neoincotinoids. It has reduced their
> longevity, behavior, etc.

3.  The Niwot Ridge litter and woody debris dataset was collected from
    the National Ecological Observatory Network, which collectively
    includes 81 aquatic and terrestrial sites across 20 ecoclimatic
    domains. 32 of these sites sample forest litter and woody debris,
    and we will focus on the Niwot Ridge long-term ecological research
    (LTER) station in Colorado. Why might we be interested in studying
    litter and woody debris that falls to the ground in forests? Feel
    free to do a brief internet search if you feel you need more
    background information.

> Answer: Litter and wood debris is important for nutrient cycling and
> moisture in forests. Addittionally, it provides habitat for
> terrestrial organisms and plays an important role in carbon
> sequestration. It shapes the structure and roughness of the forest
> ground, influencing sediment transport.

4.  How is litter and woody debris sampled as part of the NEON network?
    Read the NEON_Litterfall_UserGuide.pdf document to learn more. List
    three pieces of salient information about the sampling methods here:

> Answer: 
1. Sampling sites must contain woody vegetation that >2 meters tall.
2. Depending on the vegetation, trap placement within plots are either random
or not random. 
3. Ground traps are sampled once a year. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Dimensions}

# Finding the dimensions using dim() function
dim(Neonics)

```

6.  Using the `summary` function on the "Effect" column, determine the
    most common effects that are studied. Why might these effects
    specifically be of interest?

```{r Dimensions on Effects}

# Finding the dimensions on Effect
summary(Neonics$Effect)

```

> Answer: Mortality, population, and behavior seem to be the most common
> effects that are studied. Neonicotinoids have been shown to be toxic
> to target and non-target organisms. By understanding the phsyioloigcal
> and neural effects on these species, it might lead scientists and
> interested parties to adjust the amount of insecticide used during
> application or switch to a new insecticide.

7.  Using the `summary` function, determine the six most commonly
    studied species in the dataset (common name). What do these species
    have in common, and why might they be of interest over other
    insects? Feel free to do a brief internet search for more
    information if needed.[TIP: The `sort()` command can sort the output
    of the summary command...]

```{r Dimensions on Species Common Name}

# Using the summary function to extract Species Common Name
summary(Neonics$Species.Common.Name)

```

> Answer: The six most commonly studied species are Honey bees,
> Parasitic Wasps, Buff Tailed Bumblebees, Cariolan Honey Bees, Bumble
> Bees, Italian Honeybees. Bees are in an important pollinator in
> agricultural ecosystems. However, they are very susceptible to illness
> or death insecticides and are exposed to them frequently.

8.  Concentrations are always a numeric value. What is the class of
    `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Class of conc.1..Author}

# Using the class() function to determine Conc.1..Author class
class(Neonics$Conc.1..Author.)
```

> Answer: Conc.1..Author are factors and not numeric. Conc.1..Author is a string
that is represnting that value, so that means that Conc.1..Author has multiple
categories.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies
    conducted by publication year.

```{r Creating a Frequency line graph}

# Creating ggplot first to then use geom_freqpoly
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 30) 

```

10. Reproduce the same graph but now add a color aesthetic so that
    different Test.Location are displayed as different colors.

```{r Creating a frequency line graph and differentiating by location}

# Installing ggplot2 as my plots were not visible 
install.packages("ggplot2")
library(ggplot2)

# Creating a ggplot and differentiating test locations by color
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 30) 


```

Interpret this graph. What are the most common test locations, and do
they differ over time?

> Answer: The two most common locations are the labs and natural fields. Initially
(between 1980 - 1990), the most common test sites were in undetermine fields
and labs. But as the use of insecticide increased, it was easier for scientist
to go actual locations and conduct surveys. 

11. Create a bar graph of Endpoint counts. What are the two most common
    end points, and how are they defined? Consult the
    ECOTOX_CodeAppendix for more information.

[**TIP**: Add
`theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to
the end of your plot command to rotate and align the X-axis labels...]

```{r Creating bar graph}

# Using ggplot and then geom-bar to create a bar graph of endpoint counts
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar()
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Answer: NOEL and LOEL are the two most common points. NOEL is defined as 
no-observable-effect-level. This means that the highest concentration of insecticide
and its effect is not different from the responses of the control. LOEL is 
defined as Lowest-observable-effect-level. This indicates that the effects from a
low concentration were different from the control.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to
    a date and confirm the new class of the variable. Using the `unique`
    function, determine which dates litter was sampled in August 2018.

```{r Class of Litter$collectDate}

# Determining class of Litter$collect
class(Litter$collectDate) # it is a factor

# Changing collectDate from factor to date
Litter_CollectData <- ymd(Litter$collectDate)
Litter_CollectData

# Determining class of Litter_collectDate
class(Litter_CollectData)

# Finding dates litter were sampled in August 2018
unique(Litter_CollectData)

```

13. Using the `unique` function, determine how many plots were sampled
    at Niwot Ridge. How is the information obtained from `unique`
    different from that obtained from `summary`?

```{r Finding plots sampled at Niwot Ridge}

# Using unique () to determine the plots sampled at Niwot Ridge
unique(Litter$namedLocation) 

# From this function, it has been determined that 12 unique plots were sampled

```

> Answer: The unqiue () function is able to return a vector, data frame, etc
but removes duplicate features, returning unqiue object. The summary ()
function provides information of the raw date. It provides the returns the minimum, 
maximum, mean, median, etc. of the vector. The summary () function counts the
number of times a value appears in a certain category.

14. Create a bar graph of functionalGroup counts. This shows you what
    type of litter is collected at the Niwot Ridge sites. Notice that
    litter types are fairly equally distributed across the Niwot Ridge
    sites.

```{r Bar graph for functionalGroup Count}

# # Using ggplot and then geom-bar to create a bar graph of functional Group count
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar()


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a
    violin plot of dryMass by functionalGroup.

```{r Creating plots for drymass by functionalGroup}

# Using the geomboxplot and geom_violing functions to create plots

#geom_boxplot
ggplot(Litter) +
  geom_boxplot(aes(x = dryMass, y = functionalGroup))

#geom_violin
ggplot(Litter) +
  geom_violin(aes(x = dryMass, y = functionalGroup))

```

Why is the boxplot a more effective visualization option than the violin
plot in this case?

> Answer: The boxplot provides a better visual compared to the violin plot as it
demonstrates the minimum, median, maximum. Additionally, it shows the interquartile
range so it better demonstrates how close the dry mass weight was in distribution.
The boxplot also showed outliers in the data. The violin plot just demonstrates
the range of the dry mass.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles and twigs/branches showed to have the highest biomass. 
