---
title: "Assignment 2: Coding Basics"
author: "Tani Valdez Rivas"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basic 1}
#1.

seq(1, 30, 3) # Establishing that the sequence is starting at one, increasing by 3s until it reaches 30.
seq_to_thirty <- seq(1, 30, 3)
seq_to_thirty

#2. 

mean_seq <- mean(seq_to_thirty) # Summary statistics can help calculate the mean and median of a data set
median_seq <- median(seq_to_thirty)
mean_seq
median_seq

#3. 


mean_seq > median_seq # Asking if the mean is greater than the median


```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#5. 

Students <- c('Siobhan', 'Kendall', 'Logan', 'Tom') # The c function can be used to create a vector
Students

Scores <- c(85, 72, 91, 10) # The c function can also be used for numbers
Scores

Scores > 50 # Which scores are above 50

#7. 

# Assessing if each vector is a data frame
is.data.frame(Students)
is.data.frame(Scores)

# Converting the objects to a data frame
df_Students <- as.data.frame(Students)
df_Students
df_Scores <- as.data.frame(Scores)
df_Scores

# Binding the data together
df_Students_scores <- cbind(df_Students,df_Scores, Scores > 50)
class(df_Students_scores) # Assessing if the object is a data frame
df_Students_scores

# Labeling the column
colnames(df_Students_scores)
Label_Columns_for_Student_Scores <- data.frame("Student"=Students, "Scored"=Scores, "Passed?" = Scores > 50)
Label_Columns_for_Student_Scores

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A matrix is a data structure that is two-dimesnioal as it only contains rows and columns. Matrices possesses the same data type and has a fixed number of rows and columns. A data frame is another data structure that can contain various types of data and is one-dimensional 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement.

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}

Student_Scores <- function(Scores){ # Writing the function
  ifelse(Scores > 50, print("True"), print("False")) # Setting the return value of the condition
}

Student_Passed <- Student_Scores(Scores) # Determines if students received a passing grade
Student_Passed
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Using the 'ifelse' statement worked best to determine if a student passed with a score of 50 and above. Using the 'if' and 'else' statement required a lot more input from the user. In comparison, the 'ifelse' uses one line of code and it streamlines the process more efficiently.

